---
title: "scHOT analysis of Mouse Olfactory Bulb data"
author: "Shila Ghazanfar"
date: "05/03/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 16 
                     fig_height: 14
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE,cache.lazy = FALSE)
```

# Set up

```{r}
source("../functions.R")
tol12qualitative=c("#332288", "#6699CC", "#88CCEE", "#44AA99", "#117733", "#999933", "#DDCC77", "#661100", "#CC6677", "#AA4466", "#882255", "#AA4499")
if (!file.exists("output")) {
  system("mkdir output")
}
parallel = TRUE
ncores = 10
```

# Load packages

```{r}
if (!require(DCARS)) {
  library(devtools)
  devtools::install_github("shazanfar/DCARS")
}
library(DCARS)
library(ggplot2)
library(gplots)
library(SingleCellExperiment)
library(scater)
library(scran)
library(reshape)
library(scattermore)
library(dynamicTreeCut)
library(ComplexHeatmap)
library(GO.db)
library(org.Mm.eg.db)
library(stringr)
library(matrixStats)
library(ggforce)
library(patchwork)
library(ggpubr)
library(cowplot)
library(parallel)
library(GGally)
library(corrplot)
library(UpSetR)
```

# Process MOB data from URL

```{r}
if (!file.exists("output/counts_coords.RData")) {
  
  # data file URL https://www.spatialresearch.org/wp-content/uploads/2016/07/Rep11_MOB_count_matrix-1.tsv
  # might take a minute to download
  counts_raw = read.delim(url("https://www.spatialresearch.org/wp-content/uploads/2016/07/Rep11_MOB_count_matrix-1.tsv"), header = TRUE, row.names = 1)
  # counts_raw = read.delim(file = "Rep11_MOB_count_matrix-1.tsv", header = TRUE, row.names = 1)
  dim(counts_raw)
  
  coords_raw = do.call(rbind, strsplit(rownames(counts_raw), "x"))
  # xy plane only
  
  coords = apply(coords_raw, 1:2, as.numeric)
  colnames(coords) <- c("x","y")
  rownames(coords) <- rownames(counts_raw)
  
  plot(coords[,"x"], coords[,"y"])
  
  counts = t(counts_raw)
  
  sce = SingleCellExperiment(assays = list(counts = counts),
                             colData = coords)
  sce <- scater::normalize(sce)
  
  var.fit <- trendVar(sce, parametric=TRUE, loess.args=list(span=0.3), use.spikes = FALSE)
  var.out <- decomposeVar(sce, var.fit)
  
  pdf(file = "output/HVG_selection.pdf", height = 8, width = 8)
  plot(var.out$mean, var.out$total, pch=16, cex=0.6, xlab="Mean log-expression", 
       ylab="Variance of log-expression")
  curve(var.fit$trend(x), col="dodgerblue", lwd=2, add=TRUE)
  
  seqvals = seq(min(var.out$mean), max(var.out$mean), length.out = 1000)
  peakExp = seqvals[which.max(var.fit$trend(seqvals))]
  
  hvg.out <- var.out[which(var.out$FDR <= 0.05 & var.out$mean > peakExp),]
  nrow(hvg.out)
  hvg.out <- hvg.out[order(hvg.out$bio, decreasing=TRUE),] 
  points(var.out$mean[which(var.out$FDR <= 0.05 & var.out$mean > peakExp)], 
         var.out$total[which(var.out$FDR <= 0.05 & var.out$mean > peakExp)], col="red", pch=16)
  abline(v = peakExp, lty = 2, col = "black")
  dev.off()
  
  head(hvg.out)
  
  HVG = sort(rownames(hvg.out))
  
  expr = logcounts(sce)
  
  save(counts, coords, sce, HVG, expr, file = "output/counts_coords.RData")
} else {
  load("output/counts_coords.RData")
}

W = weightMatrix_nD(coords, span = 0.05)
```

# Test for spatial differential expression and remove

This takes a long time (approx 2 hours) to run.

```{r}
#### test for spatial differential expression by using weightedMean
# results in a list of non-spatially DE genes
samepairs = cbind(rownames(expr), rownames(expr))
rownames(samepairs) = samepairs[,1]

if (!file.exists("output/DE_stats_all.Rds")) {
  DE_stats_all = DCARSacrossNetwork(expr,
                                    samepairs,
                                    W = W,
                                    weightedConcordanceFunction = weightedMeanMatrixStats,
                                    extractTestStatisticOnly = TRUE,
                                    niter = 1000,
                                    verbose = FALSE)
  saveRDS(DE_stats_all, file = "output/DE_stats_all.Rds")
} else {
  DE_stats_all = readRDS("output/DE_stats_all.Rds")
}

if (!file.exists("output/DE_sampled_permstats.Rdata")) {
  # Does globalCor relate to the null distribution for genes?
  set.seed(500)
  globalMean = rowMeans(expr)
  pairs_sampled = samepairs[DCARS::stratifiedSample(globalMean, length = 100),]
  
  sampled_permstats = DCARSacrossNetwork(expr,
                                         pairs_sampled,
                                         W = W,
                                         weightedConcordanceFunction = weightedMeanMatrixStats,
                                         extractPermutationTestStatistics = TRUE,
                                         niter = 1000,
                                         verbose = TRUE)
  save(sampled_permstats, pairs_sampled, globalMean, file = "output/DE_sampled_permstats.Rdata")
} else {
  load("output/DE_sampled_permstats.Rdata")
}
plot(globalMean[rownames(pairs_sampled)], unlist(lapply(unlist(sampled_permstats, recursive = FALSE),quantile, 0.99)))
points(globalMean[rownames(pairs_sampled)], unlist(lapply(unlist(sampled_permstats, recursive = FALSE),quantile, 0.95)), col = "red")
points(globalMean[rownames(pairs_sampled)], unlist(lapply(unlist(sampled_permstats, recursive = FALSE),quantile, 0.90)), col = "blue")
# looks like there is a general relationship between global mean and the null distribution, but this does not appear to be a strong enough association to predict null distribution by just globalMean
# will need to run full permutation testing

if (!file.exists("output/DE_n100_permstats.Rds")) {
  set.seed(500)
  if (!parallel) {
    n100_permstats = DCARSacrossNetwork(expr,
                                        samepairs,
                                        W = W,
                                        weightedConcordanceFunction = weightedMeanMatrixStats,
                                        extractPermutationTestStatistics = TRUE,
                                        niter = 100,
                                        verbose = TRUE)
  } else {
    split_df = split.data.frame(samepairs, rep(1:ncores, length.out = nrow(samepairs)))
    names(split_df) <- NULL
    n100_permstats = mclapply(split_df, function(s) {
      res = DCARSacrossNetwork(expr,
                               s,
                               W = W,
                               weightedConcordanceFunction = weightedMeanMatrixStats,
                               extractPermutationTestStatistics = TRUE,
                               niter = 100,
                               verbose = TRUE)
      res = lapply(res, unlist)
      names(res) <- rownames(s)
      return(res)
    })
    n100_permstats = unlist(n100_permstats, recursive = FALSE)
    n100_permstats <- n100_permstats[rownames(samepairs)]
  }
  saveRDS(n100_permstats, file = "output/DE_n100_permstats.Rds")
} else {
  n100_permstats = readRDS("output/DE_n100_permstats.Rds")
}

n100_pval = sapply(names(DE_stats_all), function(gene){
  mean(unlist(n100_permstats[[gene]]) >= DE_stats_all[gene])
})

# recalculate p-value for those with pval < 0.1 with 1000 iter
# aim is to remove these genes at p-value < 0.05 level

if (!file.exists("output/DE_n1000_pval.Rds")) {
  set.seed(500)
  if (!parallel) {
    n1000_pval = DCARSacrossNetwork(expr,
                                   samepairs[n100_pval < 0.1,],
                                   W = W,
                                   weightedConcordanceFunction = weightedMeanMatrixStats,
                                   niter = 1000,
                                   verbose = TRUE)
  } else {
    split_df = split.data.frame(samepairs[n100_pval < 0.1,], rep(1:ncores, length.out = nrow(samepairs[n100_pval < 0.1,])))
    names(split_df) <- NULL
    n1000_pval = mclapply(split_df, function(s) {
      res = DCARSacrossNetwork(expr,
                               s,
                               W = W,
                               weightedConcordanceFunction = weightedMeanMatrixStats,
                               niter = 1000,
                               verbose = TRUE)
      res = lapply(res, unlist)
      names(res) <- rownames(s)
      return(res)
    })
    n1000_pval = unlist(n1000_pval)
    n1000_pval <- n1000_pval[rownames(samepairs[n100_pval<0.1,])]
  } 
  saveRDS(n1000_pval, file = "output/DE_n1000_pval.Rds")
} else {
  n1000_pval = readRDS("output/DE_n1000_pval.Rds")
}
 
DE_pval = n100_pval
DE_pval[names(n1000_pval)] <- n1000_pval

DE_fdr = p.adjust(DE_pval, method = "BH")

nonDEgenes_fdr = names(which(DE_fdr > 0.05))
saveRDS(nonDEgenes_fdr, file = "output/nonDEgenes_fdr.Rds")
nonDEgenes = names(which(DE_pval > 0.05))
saveRDS(nonDEgenes, file = "output/nonDEgenes.Rds")
```

# Perform scHOT spatial correlation testing

```{r}
pairs = t(combn(intersect(HVG, nonDEgenes),2))
rownames(pairs) <- apply(pairs,1,paste0,collapse = "_")

dim(pairs)

if (!file.exists("output/sampled_permstats.Rdata")) {
  set.seed(500)
  # Does globalCor relate to the null distribution for genes?
  globalCor = apply(pairs, 1, function(x) cor(counts[x[1],], counts[x[2],], method = "spearman"))
  pairs_sampled = pairs[DCARS::stratifiedSample(globalCor, length = 200),]
  
  if (!parallel) {
    sampled_permstats = DCARSacrossNetwork(counts,
                                           pairs_sampled,
                                           W = W,
                                           weightedConcordanceFunction = weightedSpearman,
                                           extractPermutationTestStatistics = TRUE,
                                           niter = 1000,
                                           verbose = TRUE)
  } else {
    split_df = split.data.frame(pairs_sampled, rep(1:ncores, length.out = nrow(pairs_sampled)))
    names(split_df) <- NULL
    sampled_permstats = mclapply(split_df, function(s) {
      res = DCARSacrossNetwork(counts,
                               s,
                               W = W,
                               weightedConcordanceFunction = weightedSpearman,
                               extractPermutationTestStatistics = TRUE,
                               niter = 1000,
                               verbose = TRUE)
      res = lapply(res, unlist)
      names(res) <- rownames(s)
      return(res)
    })
    sampled_permstats = unlist(sampled_permstats, recursive = FALSE)
    sampled_permstats <- sampled_permstats[rownames(pairs_sampled)]
  }
  save(sampled_permstats, pairs_sampled, globalCor, file = "output/sampled_permstats.Rdata")
} else {
  load("output/sampled_permstats.Rdata")
}

permstatsDF = data.frame(
  genepair = rep(names(sampled_permstats), times = unlist(lapply(sampled_permstats, function(x) length(unlist(x))))), 
  stat = unlist(sampled_permstats)
)
permstatsDF$globalCor = globalCor[as.character(permstatsDF$genepair)]

df_99 = data.frame(
  genepair = names(sampled_permstats),
  globalCor = globalCor[names(sampled_permstats)],
  stat_999 = unlist(lapply(sampled_permstats, function(x) quantile(unlist(x), 0.999))),
  stat_99 = unlist(lapply(sampled_permstats, function(x) quantile(unlist(x), 0.99))),
  stat_95 = unlist(lapply(sampled_permstats, function(x) quantile(unlist(x), 0.95))),
  stat_90 = unlist(lapply(sampled_permstats, function(x) quantile(unlist(x), 0.90)))
)
df_99$fitted_999 = loess(stat_999 ~ globalCor, data = df_99)$fitted
df_99$fitted_99 = loess(stat_99 ~ globalCor, data = df_99)$fitted
df_99$fitted_95 = loess(stat_95 ~ globalCor, data = df_99)$fitted
df_99$fitted_90 = loess(stat_90 ~ globalCor, data = df_99)$fitted

g = ggplot(permstatsDF, aes(x = globalCor, y = stat)) + 
  theme_classic() +
  geom_scattermore() +
  geom_point(aes(y = stat_999, colour = "0.999"), data = df_99) + 
  geom_line(aes(y = fitted_999, colour = "0.999"), data = df_99) +
  geom_point(aes(y = stat_99, colour = "0.99"), data = df_99) + 
  geom_line(aes(y = fitted_99, colour = "0.99"), data = df_99) +
  geom_point(aes(y = stat_95, colour = "0.95"), data = df_99) + 
  geom_line(aes(y = fitted_95, colour = "0.95"), data = df_99) +
  geom_point(aes(y = stat_90, colour = "0.90"), data = df_99) + 
  geom_line(aes(y = fitted_90, colour = "0.90"), data = df_99) + 
  scale_colour_manual(values = c("0.999" = tol12qualitative[4],
                                 "0.99" = tol12qualitative[3],
                                 "0.95" = tol12qualitative[2],
                                 "0.90" = tol12qualitative[1])) +
  labs(colour = "Quantile") +
  theme(panel.grid = element_blank()) +
  xlab("Global Correlation") +
  ylab("Permuted test statistics") +
  theme(legend.position = "bottom")
g
ggsave(g, file = "output/stats_globalCor_2d.pdf", height = 8, width = 8)


if (!file.exists("output/2D_p_all.Rdata")) {
  # estimate the set of p-values
  stats_all = DCARSacrossNetwork(counts,
                                 pairs,
                                 W = W,
                                 weightedConcordanceFunction = weightedSpearman,
                                 extractTestStatisticOnly = TRUE,
                                 niter = 1000,
                                 verbose = TRUE)
  
  wcors_all = t(DCARSacrossNetwork(counts,
                                   pairs,
                                   W = W,
                                   weightedConcordanceFunction = weightedSpearman,
                                   extractWcorSequenceOnly = TRUE,
                                   niter = 1000,
                                   verbose = TRUE))
  
  p_all = estimatePvaluesSpearman(stats_all, 
                                  globalCor, 
                                  sampled_permstats,
                                  usenperm = TRUE,
                                  nperm = 10000,
                                  plot = FALSE,
                                  maxDist = 2,
                                  verbose = TRUE)
  p_all$fdr <- p.adjust(p_all$pval, method = "BH")
  
  save(p_all, stats_all, wcors_all, file = "output/2D_p_all.Rdata")
} else {
  load("output/2D_p_all.Rdata")
}

p_all$gene1 = unlist(lapply(strsplit(as.character(p_all$genepair), "_"), "[", 1))
p_all$gene2 = unlist(lapply(strsplit(as.character(p_all$genepair), "_"), "[", 2))

# differentially expressed genes to remove
p_all$spatiallyDE = ifelse(as.character(p_all$gene1) %in% nonDEgenes & as.character(p_all$gene2) %in% nonDEgenes,"No", "DE")

dim(subset(p_all, spatiallyDE == "No"))
nonDEfdr = p_all$pval
nonDEfdr[p_all$spatiallyDE == "DE"] <- NA
nonDEfdr = p.adjust(nonDEfdr, method = "BH")
nonDEfdr[is.na(nonDEfdr)] <- 1
p_all$nonDEfdr = nonDEfdr


getDF = function(gene1, gene2 = NULL) {
  if (length(gene1) > 1) {
    gene2 = gene1[2]
    gene1 = gene1[1]
  }
  
  gpair = paste0(gene1,"_",gene2)
  if (!gpair %in% rownames(wcors_all)) {
    gpair = paste0(gene2,"_",gene1)
  }
  
  wcor = wcors_all[gpair,]
  
  df_res = data.frame(x = coords[,"x"], 
                      y = coords[,"y"], 
                      g1 = expr[gene1,], 
                      g2 = expr[gene2,],
                      wcor = wcor, 
                      W_min = W[which.min(wcor),], 
                      W_max = W[which.max(wcor),])
  return(df_res)
}

basicplotFunction = function(gene1, gene2 = NULL) {
  
  require(ggforce)
  require(patchwork)
  require(ggpubr)
  
  if (length(gene1) > 1) {
    gene2 = gene1[2]
    gene1 = gene1[1]
  }
  
  df_res = getDF(gene1, gene2)
  
  t = theme(legend.key.width = unit(0.5, "inches")) +
    theme(plot.title = element_text(size = 20)) +
    theme(axis.title = element_text(size = 15))
  
  g_gene1 = ggplot(df_res, aes(x = -x, y = y)) + 
    geom_point(aes(colour = g1), size = 5) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    theme(axis.text = element_blank()) +
    xlab("") +
    ylab("") +
    ggtitle(gene1) +
    labs(colour = "") +
    theme(legend.position = "bottom") +
    theme(plot.title = element_text(hjust = 0.5, face = "italic")) +
    
    scale_color_viridis_c(breaks = c(0,max(df_res$g1)),
                          limits = c(0,max(df_res$g1)),
                          labels = c("Low","High")) +
    
    t +
    coord_fixed() +
    guides(colour = guide_colourbar(title.position = "top",
                                    title.hjust = 0.5)) +
    theme(legend.title=element_text(size=15)) +
    labs(colour = "Expression") +
    NULL
  
  g_gene2 = ggplot(df_res, aes(x = -x, y = y)) + 
    geom_point(aes(colour = g2), size = 5) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    theme(axis.text = element_blank()) +
    xlab("") +
    ylab("") +
    ggtitle(gene2) +
    theme(plot.title = element_text(hjust = 0.5, face = "italic")) +
    labs(colour = "") +
    theme(legend.position = "bottom") +
    scale_color_viridis_c(breaks = c(0,max(df_res$g1)),
                          limits = c(0,max(df_res$g1)),
                          labels = c("Low","High")) +
    t +
    coord_fixed() +
    theme(legend.position = "none") +
    NULL
  
  g2 = ggplot(df_res, aes(x = -x, y = y, fill = wcor)) + 
    geom_voronoi_tile(max.radius = 1) +
    theme_minimal() + 
    theme(panel.grid = element_blank()) +
    theme(axis.text = element_blank()) +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "bottom") +
    labs(colour = "",fill = "") +
    labs(colour = "Local correlation",fill = "Local correlation") +
    ylab("") +
    xlab("") +
    ggtitle("Correlation of both genes") +
    scale_alpha_continuous(range = c(0,0.5)) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", limits = c(-1,1)) + 
    t +
    coord_fixed() +
    guides(fill = guide_colourbar(title.position = "top",
                                  title.hjust = 0.5)) +
    theme(legend.title=element_text(size=15)) +
    NULL

  g_gene1_leg = as_ggplot(get_legend(g_gene1))
  g2_leg = as_ggplot(get_legend(g2))
  
  scater::multiplot(g_gene1 + theme(legend.position = "none") + 
                      theme(plot.margin = margin(10,0,-10,0)),
                    g_gene2 + 
                      theme(plot.margin = margin(10,0,-10,0)),
                    g2 + theme(legend.position = "none") + 
                      theme(plot.margin = margin(10,0,-10,0)),
                    g_gene1_leg,
                    g2_leg,
                    layout = matrix(
                      c(1,1,2,2,3,3,
                        1,1,2,2,3,3,
                        1,1,2,2,3,3,
                        6,4,4,6,5,5), ncol = 6, byrow = TRUE))
}

plotFunction = function(gene1, gene2 = NULL) {
  
  require(ggforce)
  require(patchwork)
  
  if (length(gene1) > 1) {
    gene2 = gene1[2]
    gene1 = gene1[1]
  }
  
  df_res = getDF(gene1, gene2)
  
  g_W_min = ggplot(df_res, aes(x = g1, y = g2)) + 
    geom_point(aes(alpha = W_min, size = W_min), colour = "purple") +
    theme_minimal() +
    xlab(gene1) +
    ylab(gene2) +
    ggtitle("Min") +
    NULL
  g_W_max = ggplot(df_res, aes(x = g1, y = g2)) + 
    geom_point(aes(alpha = W_max, size = W_max), colour = "orange") +
    theme_minimal() +
    xlab(gene1) +
    ylab(gene2) +
    ggtitle("Max") +
    NULL
  
  g_xy = ggplot(df_res, aes(x = x, y = y)) + 
    geom_point(size = 0.1) +
    geom_density_2d(data = subset(df_res, W_max != 0), colour = "orange") +
    geom_density_2d(data = subset(df_res, W_min != 0), colour = "purple") +
    theme_minimal() +
    xlab("x coordinate") +
    ylab("y coordinate") +
    ggtitle("Positions") +
    NULL
  
  g_gene1 = ggplot(df_res, aes(x = x, y = y)) + 
    geom_point(aes(colour = g1), size = 5) +
    
    theme_minimal() +
    ggtitle(gene1) +
    scale_color_gradient2(low = "black", mid = "yellow", high = "red", midpoint = 2) +
    NULL
  
  g_gene2 = ggplot(df_res, aes(x = x, y = y)) + 
    geom_point(aes(colour = g2), size = 5) +
    
    theme_minimal() +
    ggtitle(gene2) +
    scale_color_gradient2(low = "black", mid = "yellow", high = "red", midpoint = 2) +
    NULL
  
  g2 = ggplot(df_res, aes(x = x, y = y, fill = wcor)) + 
    geom_voronoi_tile(max.radius = 1) +
    geom_point(size = 1, colour = "black") +
    theme_minimal() + 
    scale_alpha_continuous(range = c(0,0.5)) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", limits = c(-1,1)) + 
    NULL
  
  
  return(g_gene1 + g_gene2 + g2 + 
           g_W_min + g_W_max + g_xy + plot_layout(ncol = 3, nrow = 2, byrow = TRUE))
}

FDR_level = 0.2

wcorsSig = wcors_all[p_all$nonDEfdr < FDR_level,]
pairsSig = pairs[p_all$nonDEfdr < FDR_level,]

dim(wcorsSig)

# basic plot functions of sig pairs
apply(pairsSig,1, function(p) {
  #print(p)
  if (!file.exists("output/basicplots")) {
    system("mkdir output/basicplots")
  }
  pdf(paste0("output/basicplots/", p[1], "_", p[2], ".pdf"),
      height = 4.5, width = 12,onefile=FALSE)
  basicplotFunction(p)
  dev.off()
})
```

# Interrogate significant gene pairs

```{r}
cellsCut = cutree(hclust(cordist(t(wcorsSig))), 20)
table(cellsCut)
sort(table(cellsCut))

df_res = getDF(pairs[1,])

df_res$cellsCut <- factor(cellsCut)
ggplot(df_res, aes(x = x, y = y, fill = cellsCut, colour = cellsCut)) + 
  geom_point(size = 7, shape = 21, stroke = 1.5, colour = "black") +
  theme_minimal() + 
  scale_alpha_continuous(range = c(0,0.5)) +
  NULL
ggsave(file = "output/cellsCluster_spatial.pdf", height = 8, width = 10)

hc = hclust(dist(wcorsSig, method = "maximum"), method = "complete")

genepairsClustDynamic = cutreeDynamic(
  hc, 
  minClusterSize = 10, 
  method = "tree",
  deepSplit = TRUE,
  useMedoids = FALSE
)
kk = length(unique(genepairsClustDynamic))-1
kk
genepairsClust = cutree(hc, k = kk)
plot(hc)
table(genepairsClust)
sort(table(genepairsClust))

p_sig = cbind(p_all[names(genepairsClust),], cluster = genepairsClust)
saveRDS(p_sig, file = "output/p_sig.Rds")

write.table(as.matrix(sort_df(p_sig, c("cluster","genepair"))),
            file = "output/sig_genepairs.tsv", row.names = FALSE,
            col.names = TRUE, quote = FALSE, sep = "\t")

write.table(as.matrix(genepairsClust), file = "output/genepairsClust.tsv", row.names = TRUE,
            col.names = FALSE, quote = FALSE, sep = "\t")

genepairs_split = lapply(split(names(genepairsClust), genepairsClust), function(x) t(do.call(cbind, strsplit(x, "_"))))
genepairs_split_genes = lapply(genepairs_split, function(x) sort(unique(c(x))))

sapply(names(genepairs_split_genes), function(i){
  write(genepairs_split_genes[[i]], file = paste0("output/cluster_", i,".txt"))
})

geneClustMembers = sapply(unique(genepairsClust), function(x) 
  unique(c(pairsSig[genepairsClust == x,])), simplify = FALSE)
names(geneClustMembers) <- paste0("cluster_", unique(genepairsClust))
saveRDS(geneClustMembers, file = "output/geneClustMembers.Rds")

jacDist_pairs = expand.grid(names(geneClustMembers),names(geneClustMembers))
jacDist_vals = apply(jacDist_pairs,1,function(x){
  if (x[1] == x[2]) return(NA)
  set1 = geneClustMembers[[x[1]]]
  set2 = geneClustMembers[[x[2]]]
  return(length(intersect(set1,set2))/length(union(set1,set2)))
})
jacDist = as.matrix(reshape::cast(cbind(jacDist_pairs, jacDist_vals), formula = Var2 ~ Var1, value = "jacDist_vals"))

pdf("output/jacDist_clusters.pdf", height = 8, width = 8)
heatmap.2(jacDist, trace = "n", main = "Jaccard distance of genes within clusters", 
          density.info = "none",
          key.title = "",
          key.xlab = "Jaccard distance",
          symm = TRUE, 
          revC = TRUE)
dev.off()

meanwcorsSig = apply(wcorsSig, 2, function(x) {
  tapply(x, genepairsClust, mean)
})
rownames(meanwcorsSig) <- paste0("cluster_", 1:length(unique(genepairsClust)))

pdf("output/meanGenes_heatmap.pdf", height = 12, width = 24)
heatmap.2(meanwcorsSig, trace = "n", col = colorRampPalette(c("blue","white","red")),
          key.title = "",
          key.xlab = "Mean weighted correlation",
          main = "Mean weighted correlation of clustered genepairs")
dev.off()

df_res2 <- getDF(pairs[1,])
df_res2 <- cbind(df_res2, t(meanwcorsSig))

gList_cells <- sapply(rownames(meanwcorsSig), function(name) {
  ggplot(df_res2, aes(x = -x, y = y, fill = get(name))) + 
    geom_voronoi_tile(max.radius = 1) +
    geom_point(size = 0.5, colour = "black") +
    theme_minimal() + 
    scale_alpha_continuous(range = c(0,0.5)) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", limits = c(-1,1)) + 
    ggtitle("") +
    theme(legend.position = "none") +
    theme(panel.grid = element_blank()) +
    theme(axis.ticks = element_blank()) +
    theme(axis.text = element_blank()) +
    xlab("") + ylab("") +
    theme(plot.title = element_text(hjust = 0.5)) +
    coord_fixed() +
    NULL
}, simplify = FALSE)

gAll = patchwork::wrap_plots(gList_cells, ncol = length(gList_cells)/5, nrow = 5)
gAll
ggsave(gAll, file = "output/genepairsClust_wcors.pdf", height = 5, width = 44)

hc_mean_cells = hclust(dist(t(wcorsSig), method = "euclidean"), method = "complete")

hc_mean_cells_groups = cutreeDynamic(
  hc_mean_cells, 
  minClusterSize = 10, 
  method = "tree",
  deepSplit = FALSE,
  useMedoids = FALSE
)

kk_cells = length(unique(hc_mean_cells_groups))-1
kk_cells
hc_mean_cells_groups = cutree(hc_mean_cells, kk_cells)
plot(hc_mean_cells)
plot(hc_mean_cells, labels = hc_mean_cells_groups)
hc_mean_cells_fac = factor(hc_mean_cells_groups, levels = unique(hc_mean_cells_groups[hc_mean_cells$order]))
df_hc_mean_cells = data.frame(
  x = coords[,1],
  y = coords[,2],
  hc_mean_cells = hc_mean_cells_fac
)
s = 1
g_cells = ggplot(df_hc_mean_cells, aes(x = -x, y = y)) + 
  geom_point(data = df_hc_mean_cells[,1:2], alpha = 0.2, stroke = 0, size = s, pch = 16) + 
  geom_point(stroke = 0, size = s, pch = 16) +
  facet_grid(hc_mean_cells~.) + 
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_blank()) +
  xlab("") +
  ylab("") +
  coord_fixed() +
  NULL
g_cells
ggsave(g_cells, file = "output/split_heatmap_cells.pdf",height = 7, width = 1.5)

hc_mean_genepairs = hc
hc_mean_genepairs_groups = genepairsClust

plot(hc_mean_genepairs)
plot(hc_mean_genepairs, labels = hc_mean_genepairs_groups)
hc_mean_genepairs_fac = factor(hc_mean_genepairs_groups, levels = unique(hc_mean_genepairs_groups[hc_mean_genepairs$order]))

split(names(hc_mean_genepairs_fac), hc_mean_genepairs_fac)

write.table(sort_df(data.frame(grouping = hc_mean_genepairs_fac, cluster = names(hc_mean_genepairs_fac)), "grouping"), file = "output/hc_mean_genepairs_fac_split.txt", col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")

hc_mean_meanwcorsSig = apply(wcorsSig, 2, function(x)
  tapply(x, hc_mean_genepairs_fac, mean)
)

df_hc_mean_genepairs = data.frame(
  x = rep(coords[,1],times = nrow(hc_mean_meanwcorsSig)),
  y = rep(coords[,2],times = nrow(hc_mean_meanwcorsSig)),
  hc_mean_genepairs = factor(rep(levels(hc_mean_genepairs_fac), each = ncol(meanwcorsSig)),
                             levels = levels(hc_mean_genepairs_fac)),
  hc_mean_genepairs_wc = c(t(hc_mean_meanwcorsSig))
)

g_mean = ggplot(df_hc_mean_genepairs, aes(x = -x, y = y, colour = hc_mean_genepairs_wc)) + 
  geom_point(size = 0.7) +
  facet_grid(~hc_mean_genepairs) + 
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_blank()) +
  scale_colour_gradient2(low = "blue", mid = "white", high = "red", limits = c(-1,1)) + 
  theme(legend.position = "none") +
  xlab("") +
  ylab("") +
  coord_fixed() +
  NULL
g_mean
cowplot::ggsave2(g_mean, file = "output/split_heatmap_genepairs.pdf", height = 1.3, width = 9)

pdf("output/split_heatmap.pdf",height = 10, width = 15)
tmat = t(wcorsSig)
colnames(tmat) <- gsub("cluster_","", colnames(tmat))
h = ComplexHeatmap::Heatmap(tmat, 
                            cluster_columns = hc_mean_genepairs,
                            cluster_rows = hc_mean_cells,
                            row_dend_reorder = FALSE,
                            column_dend_reorder = FALSE,
                            row_split = kk_cells,
                            column_split = kk,
                            show_heatmap_legend = FALSE,
                            show_column_names = FALSE
)
print(h)
dev.off()
grob = grid.grabExpr(draw(h))

tmatmean = t(apply(tmat, 1, function(x) tapply(x, hc_mean_genepairs_fac, mean)))

hh = ComplexHeatmap::Heatmap(tmatmean, cluster_columns = FALSE,
                             cluster_rows = hc_mean_cells,
                             row_dend_reorder = FALSE,
                             column_dend_reorder = FALSE,
                             row_split = kk_cells,
                             show_heatmap_legend = FALSE,
                             show_column_names = FALSE,
                             column_title = "Gene pair clusters",
                             row_title = "Cells",
                             row_title_gp = gpar(fontsize = 20),
                             column_title_gp = gpar(fontsize = 20),
                             border = "black"
)
hh
grobh = grid.grabExpr(draw(hh))

pdf("output/split_heatmap_combined.pdf",height = 7, width = 11, useDingbats = FALSE)
cowplot::plot_grid(grob, 
                   g_cells + theme(plot.margin = margin(2,0.5,0.2,-0.5,unit = "cm")) +
                     theme(strip.text = element_blank()), 
                   g_mean + theme(plot.margin = margin(0,0,0,1.5, unit = "cm")) +
                     theme(strip.text = element_blank()) +
                     NULL
                   , 
                   ncol = 2,
                   rel_heights = c(7,1),
                   rel_widths = c(9,1))
dev.off()

pdf("output/split_heatmap_combined_summarised.pdf",height = 8, width = 9, useDingbats = FALSE)
cowplot::plot_grid(grobh, 
                   g_cells + 
                     theme(plot.margin = margin(1.3,0.5,-0.2,-0.5,unit = "cm")) +
                     theme(strip.text = element_blank()), 
                   g_mean + 
                     theme(plot.margin = margin(0,0,0,1.5, unit = "cm")) +
                     theme(strip.text = element_blank()), 
                   ncol = 2,
                   rel_heights = c(8,1),
                   rel_widths = c(8,1))

dev.off()
```

# Gene ontology enrichment testing

```{r}
if (!file.exists("output/GO_list.Rds")) {
  library(GO.db)
  library(org.Mm.eg.db)
  keys = keys(org.Mm.eg.db)
  columns(org.Mm.eg.db)
  GO_info = select(org.Mm.eg.db, keys=keys, columns = c("SYMBOL", "GO"))
  
  keep = GO_info$SYMBOL %in% rownames(counts)
  table(keep)
  GO_info_filt = GO_info[keep,]
  
  # at least 10 genes in the term and less than 500
  # allTerms = names(which(table(GO_info_filt$GO) >= 10 & table(GO_info_filt$GO) <= 500))
  sizeTerms = names(which(table(GO_info_filt$GO) >= 10 & table(GO_info_filt$GO) <= 500))
  
  # also have at least one of the genes which we actually test
  testTerms = subset(GO_info_filt, SYMBOL %in% intersect(HVG, nonDEgenes))$GO
  
  allTerms = intersect(sizeTerms, testTerms)
  
  GO_info_terms = select(GO.db, columns = columns(GO.db), keys = allTerms)
  rownames(GO_info_terms) <- GO_info_terms$GOID
  
  allTermNames = GO_info_terms[allTerms, "TERM"]
  names(allTermNames) <- allTerms
  
  GO_list = sapply(allTerms, function(term) {
    print(term)
    genes = GO_info_filt[GO_info_filt$GO == term, "SYMBOL"]
    return(sort(unique(genes[!is.na(genes)])))
  }, simplify = FALSE)
  names(GO_list) <- allTermNames
  
  saveRDS(GO_list, file = "output/GO_list.Rds")
} else {
  GO_list = readRDS("output/GO_list.Rds")
}

if (!file.exists("output/superclusterstotest_GO.Rds")) {
  superclusterstotest_GO = lapply(geneClustMembers, function(set) {
    print("testing..")
    genesetGOtest(set, rownames(counts), GO_list)
  }
  )
  saveRDS(superclusterstotest_GO, file = "output/superclusterstotest_GO.Rds")
} else {
  superclusterstotest_GO <- readRDS("output/superclusterstotest_GO.Rds")
}

gList = lapply(superclusterstotest_GO, function(pval) {
  df = data.frame(term = factor(names(pval), levels = c(names(pval), "")),
                  pval = pval,
                  qval = p.adjust(pval, method = "BH"))
  df$label = df$term
  df$label[pval != 1] <- ""
  df_sorted = sort_df(df, "pval")[1:10,]
  df_sorted$term = factor(df_sorted$term, levels =  rev(df_sorted$term))
  
  g = ggplot(df_sorted, aes(x = term, y = -log10(pval), fill = qval < 0.05)) +
    theme_classic() +
    geom_col() +
    coord_flip() +
    xlab("") +
    ylab("-log10(P-value)") +
    # geom_hline(yintercept = -log10(0.01), colour = "red", linetype = "dashed", size = 1) +
    scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) +
    scale_fill_manual(values = c("TRUE" = "dimgrey", "FALSE" = "peachpuff")) +
    theme(legend.position = "none") +
    NULL
  return(g)
  
})

gTogether = sapply(seq_len(length(gList)), function(i){
  
  g_1 = gList[[i]] + 
    theme(axis.text = element_text(size = 12))
  n = ggtitle(paste0("Cluster ",i)) 
  g_2 = gList_cells[[i]] + theme(title = element_text(size = 20)) + 
    coord_fixed()
  
  pdf(paste0("output/go_cluster_plot_cluster_",i,".pdf"), 
      height = 7, width = 4.5)
  scater::multiplot(g_2 + n, g_1 + ylab(""), layout = matrix(c(1,1,2,2,2), ncol = 1))
  dev.off()
  
  return(g_1 + g_2)
}, simplify = FALSE)
```

## Compare different span choices

Using script run on cluster due to high computational demand. Note that the file
span_output.RData is around 60MB.

```{r}
if (!file.exists("output/span_output.RData")) {
  source("span_MOB.R")
}
load("output/span_output.RData")

span_p_all_all = do.call(rbind, span_p_all)
span_p_all_all$testing <- rep(names(span_p_all), 
                                         times = unlist(lapply(span_p_all, nrow)))

span_pvals = do.call(cbind,lapply(span_p_all, "[", "pval"))
colnames(span_pvals) <- names(span_p_all)

span_cor = cor(-log10(span_pvals), method = "spearman")

g = ggpairs(-log10(span_pvals),
            
            lower = list(continuous = wrap("points", alpha = 0.3, size=0.1), 
                         combo = wrap("dot", alpha = 0.4, size=0.2)),
            title = "-log10(P-value)"
)

g
ggsave(g, file = "output/span_pairs.pdf", height = 12, width = 14)


pdf("output/span_corrplot.pdf", height = 10, width = 10)
g_cor = corrplot(span_cor, order = "original")
dev.off()

pdf("output/span_upset.pdf", height = 10, width = 16, onefile = FALSE)
upset(data = as.data.frame(as.matrix(1*(apply(span_pvals,2,p.adjust, method = "BH") < 0.2))),
      sets = colnames(span_pvals),
      order.by = "freq",
      text.scale = 2)
dev.off()

# distribution graphs
span_distn = data.frame(
    quantile_999 = unlist(lapply(span_perms, function(x) unlist(lapply(x, quantile, probs = 0.999)))),
    quantile_90 = unlist(lapply(span_perms, function(x) unlist(lapply(x, quantile, probs = 0.90)))),
    quantile_95 = unlist(lapply(span_perms, function(x) unlist(lapply(x, quantile, probs = 0.95)))),
    quantile_99 = unlist(lapply(span_perms, function(x) unlist(lapply(x, quantile, probs = 0.99)))),
    global = unlist(lapply(span_global, "[", span_sampled_ind)),
    span = rep(names(span_perms), times = unlist(lapply(span_perms, length)))
)
span_distn <- reshape::sort_df(span_distn, "global")

span_distn$quantile_999_fitted = unsplit(lapply(split.data.frame(span_distn, span_distn$span),
                                                           function(df){
                                                               loess(quantile_999 ~ global, data = df)$fitted
                                                           }), span_distn$span)

span_distn$quantile_90_fitted = unsplit(lapply(split.data.frame(span_distn, span_distn$span),
                                                          function(df){
                                                              loess(quantile_90 ~ global, data = df)$fitted
                                                          }), span_distn$span)

span_distn$quantile_95_fitted = unsplit(lapply(split.data.frame(span_distn, span_distn$span),
                                                          function(df){
                                                              loess(quantile_95 ~ global, data = df)$fitted
                                                          }), span_distn$span)

span_distn$quantile_99_fitted = unsplit(lapply(split.data.frame(span_distn, span_distn$span),
                                                          function(df){
                                                              loess(quantile_99 ~ global, data = df)$fitted
                                                          }), span_distn$span)

span_distn$span = factor(span_distn$span,levels = names(span_perms))

span_nicecolours = RColorBrewer::brewer.pal(name = "Blues", 9)
names(span_nicecolours) <- c(rep("", 5), "0.90","0.95","0.99","0.999")
g = ggplot(span_distn, aes(x = global))  + 
    
    geom_point(aes(y = quantile_999, colour = "0.999"),
               alpha = 0.5) +#, colour = span_nicecolours[9]) +
    geom_line(aes(y = quantile_999_fitted), colour = span_nicecolours[9]) +
    
    geom_point(aes(y = quantile_99, colour = "0.99"),
               alpha = 0.5) + #, colour = span_nicecolours[8]) +
    geom_line(aes(y = quantile_99_fitted), colour = span_nicecolours[8]) +
    
    geom_point(aes(y = quantile_95, colour = "0.95"),
               alpha = 0.5) + #, colour = span_nicecolours[7]) +
    geom_line(aes(y = quantile_95_fitted), colour = span_nicecolours[7]) +
    
    geom_point(aes(y = quantile_90, colour = "0.90"),
               alpha = 0.5) + #, colour = span_nicecolours[6]) +
    geom_line(aes(y = quantile_90_fitted), colour = span_nicecolours[6]) +
    facet_wrap(~span, scales = "free", ncol = 4) + 
    theme_classic() + 
    xlab("Global higher order statistic value") +
    ylab("Permuted scHOT test statistic quantile") +
    scale_colour_manual(name="Quantiles",values=span_nicecolours[6:9]) +
    theme(legend.position = "bottom") +
    NULL
g
ggsave(g, file = "output/span_pvalEstimation.pdf",
       height = 12, width = 14)
```

```{r}
sessionInfo()
```